# Ecosystem Stability: Mechanisms and Patterns
#### Ecosystem stability exercise and assignment for Wild6770, Plant Community Ecology (Fall 2016)

-------------------

**Complete readings by:** September 26, 2016  
**Assignemnt due:** October x, 2016

-------------------

### Readings
1. Loreau, M. and C. de Mazancourt. 2013. Biodiversity and ecosystem stability: a synthesis of underlying mechanisms. *Ecology Letters* **16**:106-115. [[pdf]](http://github.com/atredennick/EcoStability_Wild6770/pdfs/Loreau_2013_ELE.pdf) [[web]](http://onlinelibrary.wiley.com/doi/10.1111/ele.12073/abstract)

2. Bai, Y., H. Xingguo, J. Wu, Z. Chen, and L. Li. 2004. Ecosystem stability and compensatory effects in the Inner Mongolia grassland. *Nature* **431**:181-184. [[pdf]](http://github.com/atredennick/EcoStability_Wild6770/pdfs/Bai_2004_Nature.pdf) [[web]](http://www.nature.com/nature/journal/v431/n7005/abs/nature02850.html)

-------------------

## In-Class Exercise
Building on your expertise with the 2-species Lotka-Volterra model, we will use model simulations to help us understand the mechanisms that increase, or decrease, ecosystem stability in competitive communities. Recall the Lotka-Volterra equations with environmental stochasticity ($z_{i,t}$):

$$
N_{1,t+1} = N_{1,t} + N_{1,t} \left[r_1(1-\alpha_{11}N_{1,t} - \alpha_{12}N_{2,t}) + z_{1,t} \right], \\
N_{2,t+1} = N_{2,t} + N_{2,t} \left[r_2(1-\alpha_{21}N_{1,t} - \alpha_{22}N_{2,t}) + z_{2,t} \right].
$$

As before, we are using a version without carrying capacity ($K$), so this formulation will look a little bit different than the equations in Loreau and de Mazancourt (2013). But, the essential features are still there. For example, we still have inter ($\alpha_{ij}$'s) and intraspecific ($\alpha_{ii}$'s) competition and environmental stochasticity ($z_{i,t}$ in our model, $\sigma_{ei}\mu_{ei}(t)$ in Loreau and de Mazancourt's model). Therefore, we should be able to reproduce Loreau and de Mazancourt's results, at least qualitatively.

### Is competition stabilizing or destabilizing?
One of the most interesting and novel contributions in Loreau and de Mazancourt (2013) is on the role of interspecific competition in determining ecosystem stability. Working with a partner, discuss whether interspecific competition is stabilizing or destabilizing, and why, based on your reading of Loreau and de Mazancourt. Then, use the environmentally stochastic Lotka-Volterra model to prove it to yourself. A couple things might be helpful here:

1. You'll want to measure the synchrony of your simulated 2-species community. Fortunately, there is an R package for this, ```synchrony```. To get the package and calculate synchrony, follow these steps in ```R```:

```{r synch, eval=FALSE}
install.packages("synchrony")
synch <- community.sync(N)      # calculates synchrony and temporal correlation of community matrix N
```

2. You'll want to calculate the coefficient of variation of the total community biomass. A couple handy functions:

```{r handys, eval=FALSE}
rowSums()
mean()
sd()
```